---
aliases:
  - /new_pages/data_used.html
---

# Download handbook and data {#data-used}


<!-- Note to self: If you want to create a download link to Github, right-click the "View Raw" button on Github, copy the address, and use that in the HTML below. -->




## Download offline handbook

You can download the offline version of this handbook as an HTML file so that you can view the file in your web browser even if you no longer have internet access. If you are considering offline use of the Epi R Handbook here are a few things to consider:

* When you open the file it may take a minute or two for the images and the Table of Contents to load
* The offline handbook has a slightly different layout - one very long page with Table of Contents on the left. To search for specific terms use Ctrl+f (Cmd-f)
* See the [Suggested packages](packages_suggested.qmd) page to assist you with installing appropriate R packages before you lose internet connectivity

**To download the handbook:**

For quick access, **right-click** [this link](https://github.com/appliedepi/epirhandbook/raw/main/offline_long/index.html) **and select "Save link as"**.

If on a Mac, use Cmd+click. If on a mobile, press and hold the link and select "Save link". The handbook will download to your device. If a screen with raw HTML code appears, ensure you have followed the above instructions.




## Download data to follow along

To "follow along" with the handbook pages, you can load the example data straight into R with our companion R package, **appliedepidata**. It contains every example dataset used in this handbook.

### Install the package {.unnumbered}

**appliedepidata** is not on CRAN, so install it from its [Github repository](https://github.com/appliedepi/appliedepidata) with **pak**:

```{r, eval=F}
# install the latest version of the appliedepidata package
pak::pak("appliedepi/appliedepidata")
```

### Load a dataset directly {.unnumbered}

Use `get_data()` with the dataset's `name = ` to load it straight into R - no download, no file path, no `import()` step. The function returns the dataset as an R object (a data frame, an `sf` object, a `phylo` tree, etc., depending on the dataset).

```{r, eval=F}
# load the cleaned case linelist directly into R
linelist <- appliedepidata::get_data(name = "linelist_cleaned_rds")
```

Each page of this handbook that uses example data shows the `get_data(name = "...")` call needed for that page, right where the data are first used.

### Save a dataset to a real file {.unnumbered}

Some pages teach you how to import a *file* (e.g. an Excel workbook, or a folder of files) rather than an R object. For these, use `save_data()` to write the real file(s) to a folder of your choice with `name = ` and `path = `:

```{r, eval=F}
# write the raw linelist Excel file to your working directory
appliedepidata::save_data(name = "case_linelists_linelist_raw", path = getwd())
```

You can then `import()` it as you would any file on your computer - see the [Import and export](importing.qmd) page for details. Note that `save_data()` does not automatically unzip datasets that are distributed as a .zip bundle (e.g. multi-file shapefiles, or folders of files); use `utils::unzip()` on the saved .zip afterwards.

### Browse what's available {.unnumbered}

* `appliedepidata::search_data()` launches an interactive Shiny app so you can filter and search all the datasets in the package by language and keyword.
* `appliedepidata::list_data()` returns a data frame listing every dataset name programmatically, for use in scripts.

```{r, eval=F}
# browse datasets interactively
appliedepidata::search_data()

# list all dataset names programmatically
appliedepidata::list_data()
```

If you wish, you can also review the raw data files in the **["data" folder](https://github.com/appliedepi/appliedepidata)** of the **appliedepidata** Github repository.



### Dataset reference, by page {.unnumbered}

Below is the `get_data()`/`save_data()` name to use for the example data referenced on each page of this handbook.


#### Case linelist {.unnumbered}

This is a fictional Ebola outbreak, expanded by the handbook team from the `ebola_sim` practice dataset in the **outbreaks** package.

```{r, eval=F}
# the "raw" linelist - an Excel spreadsheet with messy data
# use this to follow along with the Cleaning data and core functions page
raw_linelist <- appliedepidata::get_data(name = "case_linelists_linelist_raw")

# the "clean" linelist - an R-specific .rds file that preserves column classes
# use this for all other pages of this handbook that use the linelist
linelist <- appliedepidata::get_data(name = "linelist_cleaned_rds")

# the "clean" linelist, as an Excel file instead
linelist_excel <- appliedepidata::get_data(name = "linelist_cleaned_excel")
```

Part of the cleaning page uses a "cleaning dictionary" (.csv file). You can load it directly into R by running the following command:

```{r, eval=F}
cleaning_dict <- appliedepidata::get_data(name = "case_linelists_cleaning_dict")
```


#### Malaria count data {#data_malaria .unnumbered}

These data are fictional counts of malaria cases by age group, facility, and day. A .rds file is an R-specific file type that preserves column classes. This ensures you will have only minimal cleaning to do after importing the data into R.

```{r, eval=F}
malaria_data <- appliedepidata::get_data(name = "malaria_facility_count_data")
```


#### Likert-scale data {.unnumbered}

These are fictional data from a Likert-style survey, used in the page on [Demographic pyramids and Likert-scales](age_pyramid.qmd).

```{r, eval=F}
likert_data <- appliedepidata::get_data(name = "likert_data")
```


#### Flexdashboard {.unnumbered}

Below are links to the files associated with the page on [Dashboards with R Markdown](flexdashboard.qmd). These are R Markdown/HTML source files, not example datasets, so they are not part of the **appliedepidata** package - download them directly from Github:

* To download the R Markdown for the outbreak dashboard, right-click this [link](https://github.com/appliedepi/epirhandbook/raw/main/data/flexdashboard/outbreak_dashboard.Rmd) (Cmd+click for Mac) and select "Save link as".
* To download the HTML dashboard, right-click this [link](https://github.com/appliedepi/epirhandbook/raw/main/data/flexdashboard/outbreak_dashboard_test.html) (Cmd+click for Mac) and select "Save link as".

#### Contact Tracing {.unnumbered}

The [Contact Tracing](contact_tracing.qmd) page demonstrated analysis of contact tracing data, using example data from [Go.Data](https://github.com/WorldHealthOrganization/godata/tree/master/analytics/r-reporting). This data is not part of **appliedepidata**; the data used in the page can be downloaded as .rds files by clicking the following links:

<a href='https://github.com/WorldHealthOrganization/godata/blob/master/analytics/r-reporting/data/cases_clean.rds?raw=true' class='download-button'>
	Click to download
	<span>the case investigation data (.rds file)</span>
</a>

<a href='https://github.com/WorldHealthOrganization/godata/blob/master/analytics/r-reporting/data/contacts_clean.rds?raw=true' class='download-button'>
	Click to download
	<span>the contact registration data (.rds file)</span>
</a>

<a href='https://github.com/WorldHealthOrganization/godata/blob/master/analytics/r-reporting/data/followups_clean.rds?raw=true' class='download-button'>
	Click to download
	<span>the contact follow-up data (.rds file)</span>
</a>



<span style="color: black;">**_NOTE:_** Structured contact tracing data from other software (e.g. KoBo, DHIS2 Tracker, CommCare) may look different. If you would like to contribute alternative sample data or content for this page, please [contact us](#contact_us).</span>

<span style="color: darkgreen;">**_TIP:_** If you are deploying Go.Data and want to connect to your instance's API, see the Import and export page [(API section)](#import_api) and the [Go.Data Community of Practice](https://community-godata.who.int/).</span>


#### GIS {.unnumbered}

Shapefiles have many sub-component files, each with a different file extension. One file will have the ".shp" extension, but others may have ".dbf", ".prj", etc. **appliedepidata** bundles each shapefile's component files together as a single .zip, so `save_data()` and then unzip:

```{r, eval=F}
# load the Sierra Leone admin-3 shapefile directly as an sf object
sle_adm3 <- appliedepidata::get_data(name = "sle_adm3")

# OR save the raw shapefile component files to a folder
shp_dir <- "sle_adm3_files"
dir.create(shp_dir)
appliedepidata::save_data(name = "sle_adm3", path = shp_dir)
utils::unzip(file.path(shp_dir, "sle_adm3.zip"), exdir = shp_dir)
```

See the [GIS basics](gis.qmd) page for how to import a shapefile with `st_read()` from the **sf** package. The [GIS basics](gis.qmd) page also links to the *Humanitarian Data Exchange* website, where you can download further shapefiles directly as zipped files (for example, the health facility points data can be downloaded [here](https://data.humdata.org/dataset/hotosm_sierra_leone_health_facilities)).


#### Phylogenetic trees {.unnumbered}

See the page on [Phylogenetic trees](phylogenetic_trees.qmd). Newick file of phylogenetic tree constructed from whole genome sequencing of 299 Shigella sonnei samples and corresponding sample data (converted to a text file). The Belgian samples and resulting data are kindly provided by the Belgian NRC for Salmonella and Shigella in the scope of a project conducted by an ECDC EUPHEM Fellow, and will also be published in a manuscript. The international data are openly available on public databases (ncbi) and have been previously published.

```{r, eval=F}
# the phylogenetic tree, as a "phylo" object (ape package)
tree <- appliedepidata::get_data(name = "Shigella_tree")

# additional information on each sample
sample_data <- appliedepidata::get_data(name = "sample_data_Shigella_tree")

# the subset-tree created later in the page
subtree <- appliedepidata::get_data(name = "Shigella_subtree_2")
```


#### Standardization {.unnumbered}

See the page on [Standardised rates](standardization.qmd). You can load the data directly with the following commands:

```{r, eval=F}
##############
# Country A
##############
# demographics for country A
A_demo <- appliedepidata::get_data(name = "country_demographics")

# deaths for country A
A_deaths <- appliedepidata::get_data(name = "deaths_countryA")

##############
# Country B
##############
# demographics for country B
B_demo <- appliedepidata::get_data(name = "country_demographics_2")

# deaths for country B
B_deaths <- appliedepidata::get_data(name = "deaths_countryB")


###############
# Reference Pop
###############
# world standard population, by sex
standard_pop_data <- appliedepidata::get_data(name = "world_standard_population_by_sex")
```



#### Time series and outbreak detection {#data_outbreak .unnumbered}

See the page on [Time series and outbreak detection](time_series.qmd). We use campylobacter cases reported in Germany 2002-2011, as available from the **surveillance** R package. (*nb.* this dataset has been adapted from the original, in that 3 months of data have been deleted from the end of 2011 for demonstration purposes)

```{r, eval=F}
counts <- appliedepidata::get_data(name = "campylobacter_germany")
```

We also use climate data from Germany 2002-2011 (temperature in degrees celsius and rain fall in millimetres). These were downloaded from the EU Copernicus satellite reanalysis dataset using the **ecmwfr** package, and are bundled by **appliedepidata** as one .zip of ten yearly .nc files. `get_data()` reads the whole bundle in as one combined **stars** object directly; `save_data()` gives you the raw .nc files if you want to inspect or process them yourself.

```{r, eval=F}
# read the combined weather data directly as a stars object
weather_data <- appliedepidata::get_data(name = "germany_weather")

# OR save the raw yearly .nc files to a folder
weather_dir <- "germany_weather_files"
dir.create(weather_dir)
appliedepidata::save_data(name = "germany_weather", path = weather_dir)
utils::unzip(file.path(weather_dir, "germany_weather.zip"), exdir = weather_dir)
```



#### Survey analysis {#data_survey .unnumbered}

For the [survey analysis](survey_analysis.qmd) page we use fictional mortality survey data based off MSF OCA survey templates. This fictional data was generated as part of the ["R4Epis" project](https://r4epis.netlify.app/).

```{r, eval=F}
# fictional survey data
survey_data <- appliedepidata::get_data(name = "survey_data")

# fictional survey data dictionary
survey_dict <- appliedepidata::get_data(name = "survey_dict")

# fictional survey population data
population <- appliedepidata::get_data(name = "population")
```




#### Shiny {#data_shiny .unnumbered}

The page on [Dashboards with Shiny](shiny_basics.qmd) demonstrates the construction of a simple app to display malaria data. The app's data are part of **appliedepidata**:

```{r, eval=F}
malaria_data <- appliedepidata::get_data(name = "malaria_facility_count_data")
```

The app's R scripts are not example data, so they are not part of **appliedepidata** - download them directly from Github:

You can <a href='https://github.com/appliedepi/epirhandbook/raw/main/data/malaria_app/app.R' class='download-button'>
	click here to download the app.R file<span> that contains both the UI and Server code for the Shiny app.</span></a>

You can <a href='https://github.com/appliedepi/epirhandbook/raw/main/data/malaria_app/global.R' class='download-button'>
	click here to download the global.R file<span></a> that should run prior to the app opening, as explained in the page.

You can <a href='https://github.com/appliedepi/epirhandbook/raw/main/data/malaria_app/funcs/plot_epicurve.R' class='download-button'>
	click here to download the plot_epicurve.R file<span></a> that is sourced by global.R. Note that you may need to store it within a "funcs" folder for the here() file paths to work correctly.
