---
aliases:
  - /new_pages/epidemic_models.html
---

# Epidemic modeling { }  


<!-- ======================================================= -->
## Overview {  }

There exists a growing body of tools for epidemic modelling that lets us conduct
fairly complex analyses with minimal effort. This section will provide an
overview on how to use these tools to:

* estimate the effective reproduction number R<sub>t</sub> and related statistics
  such as the doubling time
* produce short-term projections of future incidence

It is *not* intended as an overview of the methodologies and statistical methods
underlying these tools, so please refer to the Resources tab for links to some
papers covering this. Make sure you have an understanding of
the methods before using these tools; this will ensure you can accurately
interpret their results.

Below is an example of one of the outputs we'll be producing in this section.

```{r out.width=c('100%', '100%'), fig.show='hold', echo=F, fig.width = 12, fig.height = 9, message=F, warning=F}

## install and load packages
pacman::p_load(tidyverse, EpiNow2, EpiEstim, incidence2, incidence, epicontacts, projections)

## load linelist
linelist <- appliedepidata::get_data(name = "linelist_cleaned_rds")

## generate contacts
contacts <- linelist %>%
  transmute(
    from = infector,
    to = case_id
  ) %>%
  drop_na()

## generate epicontacts
epic <- make_epicontacts(
  linelist = linelist,
  contacts = contacts, 
  directed = TRUE
)

## load the stored EpiNow2 estimates and the case counts that the fit used
estimates <- appliedepidata::get_data(name = "epidemic_models_summarised_estimates")
observations <- appliedepidata::get_data(name = "epidemic_models_reported_cases")

## plot summary figure
ggplot(
  data = filter(estimates, variable %in% c("R", "reported_cases")),
  aes(x = date)
) +
  ## observed case counts, in the reported cases panel only
  geom_col(
    data = mutate(observations, variable = "reported_cases"),
    aes(y = confirm),
    fill = "grey70",
    width = 1
  ) +
  ## 90% credible interval and median, coloured by estimate type
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90, fill = type), alpha = 0.3) +
  geom_line(aes(y = median, colour = type)) +
  ## use label_parsed to allow subscript label
  facet_wrap(
    ~ variable,
    ncol = 1,
    scales = "free_y",
    labeller = as_labeller(c("R" = "R[t]", "reported_cases" = "Reported~cases"), label_parsed),
    strip.position = 'left'
  ) +
  labs(x = NULL, y = NULL, colour = NULL, fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    strip.background = element_blank(),
    strip.placement = 'outside',
    legend.position = 'bottom'
  )

```

<!-- ======================================================= -->
## Preparation {  }

We will use two different methods and packages for R<sub>t</sub> estimation,
namely **EpiNow** and **EpiEstim**, as well as the **projections** package for
forecasting case incidence.  

This code chunk shows the loading of packages required for the analyses. 
In this handbook we emphasize `p_load()` from **pacman**, which installs the package if necessary *and* loads it for use. 
You can also load installed packages with  `library()` from **base** R. See the page on [R basics](basics.qmd) for more information on R packages.  

	
```{r epidemic_models_packages, }
pacman::p_load(
   rio,          # File import
   here,         # File locator
   tidyverse,    # Data management + ggplot2 graphics
   epicontacts,  # Analysing transmission networks
   EpiNow2,      # Rt estimation
   EpiEstim,     # Rt estimation
   projections,  # Incidence projections
   incidence2,   # Handling incidence data
   incidence,    # Incidence objects for projections
   epitrix,      # Useful epi functions
   distcrete     # Discrete delay distributions
)
```
	
We will use the cleaned case linelist for all analyses in this section. If you want to follow along, <a href='https://github.com/appliedepi/epirhandbook_eng/raw/master/data/case_linelists/linelist_cleaned.rds' class='download-button'>click to download the "clean" linelist</a> (as .rds file). See the [Download handbook and data](data_used.qmd) page to download all example data used in this handbook.  

```{r, echo=F}
# import the linelist into R
linelist <- appliedepidata::get_data(name = "linelist_cleaned_rds")
```

```{r eval=F}
# import the cleaned linelist
linelist <- import("linelist_cleaned.rds")
```


<!-- ======================================================= -->
## Estimating R<sub>t</sub> {  }

### EpiNow2 vs. EpiEstim {.unnumbered}

The reproduction number R is a measure of the transmissibility of a disease and
is defined as the expected number of secondary cases per infected case. In a
fully susceptible population, this value represents the basic reproduction
number R<sub>0</sub>. However, as the number of susceptible individuals in a
population changes over the course of an outbreak or pandemic, and as various
response measures are implemented, the most commonly used measure of
transmissibility is the effective reproduction number R<sub>t</sub>; this is
defined as the expected number of secondary cases per infected case at a given
time _t_.

The **EpiNow2** package provides the most sophisticated framework for estimating
R<sub>t</sub>. It has two key advantages over the other commonly used package,
**EpiEstim**:

* It accounts for delays in reporting and can therefore estimate R<sub>t</sub>
  even when recent data is incomplete.
* It estimates R<sub>t</sub> on _dates of infection_ rather than the dates of
  onset of reporting, which means that the effect of an intervention will
  be immediately reflected in a change in R<sub>t</sub>, rather than with a
  delay.

However, it also has two key disadvantages:

* It requires knowledge of the generation time distribution (i.e. distribution
  of delays between infection of a primary and secondary cases), incubation
  period distribution (i.e. distribution of delays between infection and symptom
  onset) and any further delay distribution relevant to your data (e.g. if you
  have dates of reporting, you require the distribution of delays from symptom
  onset to reporting). While this will allow more accurate estimation of
  R<sub>t</sub>, **EpiEstim** only requires the serial interval distribution
  (i.e. the distribution of delays between symptom onset of a primary and a
  secondary case), which may be the only distribution available to you.
* **EpiNow2** is significantly slower than **EpiEstim**, anecdotally by a factor
  of about 100-1000! For example, estimating R<sub>t</sub> for the sample outbreak
  considered in this section takes about four hours (this was run for a large
  number of iterations to ensure high accuracy and could probably be reduced if
  necessary, however the points stands that the algorithm is slow in
  general). This may be unfeasible if you are regularly updating your
  R<sub>t</sub> estimates.
  
Which package you choose to use will therefore depend on the data, time and
computational resources available to you.

### EpiNow2 {.unnumbered}

#### Estimating delay distributions {.unnumbered}

The delay distributions required to run **EpiNow2** depend on the data you
have. Essentially, you need to be able to describe the delay from the date of
infection to the date of the event you want to use to estimate R<sub>t</sub>. If
you are using dates of onset, this would simply be the incubation period
distribution. If you are using dates of reporting, you require the
delay from infection to reporting. As this distribution is unlikely to be known
directly, **EpiNow2** lets you chain multiple delay distributions together; in
this case, the delay from infection to symptom onset (e.g. the incubation
period, which is likely known) and from symptom onset to reporting (which you
can often estimate from the data).

As we have the dates of onset for all our cases in the example linelist, we will
only require the incubation period distribution to link our data (e.g. dates of
symptom onset) to the date of infection. We can either estimate this distribution
from the data or use values from the literature.

A literature estimate of the incubation period of Ebola (taken
from [this paper](https://www.nejm.org/doi/full/10.1056/nejmoa1411100)) with a
mean of 9.1, standard deviation of 7.3 and maximum value of 30 would be
specified as follows:

```{r epidemic_models_incubation_literature, eval=F}
incubation_period_lit <- LogNormal(
  mean = 9.1,
  sd = 7.3,
  max = 30
)
```
`LogNormal()` takes the mean and the standard deviation in days, on the natural
scale, and converts them to the log scale itself. The `max` parameter cuts the
distribution at 30 days.

In this analysis, we instead estimate the incubation period distribution
from the linelist itself using the function `bootstrapped_dist_fit`, which will
fit a lognormal distribution to the observed delays between infection and onset
in the linelist.

```{r epidemic_models_incubation_estimate, eval=F}
## estimate incubation period
incubation_period <- bootstrapped_dist_fit(
  linelist$date_onset - linelist$date_infection,
  dist = "lognormal",
  max_value = 100,
  bootstraps = 1
)
```

The other distribution we require is the generation time. As we have data on
infection times __and__ transmission links, we can estimate this
distribution from the linelist by calculating the delay between infection times
of infector-infectee pairs. To do this, we use the handy `get_pairwise` function
from the package **epicontacts**, which allows us to calculate pairwise
differences of linelist properties between transmission pairs. We first create an
epicontacts object (see [Transmission chains](transmission_chains.qmd) page for further
details):

```{r epidemic_models_epicontacts, eval=F}
## generate contacts
contacts <- linelist %>%
  transmute(
    from = infector,
    to = case_id
  ) %>%
  drop_na()

## generate epicontacts object
epic <- make_epicontacts(
  linelist = linelist,
  contacts = contacts, 
  directed = TRUE
)
```

We then fit the difference in infection times between transmission pairs,
calculated using `get_pairwise`, to a gamma distribution:

```{r epidemic_models_generation_estimate, eval=F}
## estimate gamma generation time
generation_time <- bootstrapped_dist_fit(
  get_pairwise(epic, "date_infection"),
  dist = "gamma",
  max_value = 20,
  bootstraps = 1
)
```

#### Running **EpiNow2** {.unnumbered}

Now we just need to calculate daily incidence from the linelist, which we can do
easily with the **dplyr** functions `group_by()` and `n()`. Note
that **EpiNow2** requires the column names to  be `date` and `confirm`.
**EpiNow2** also stops with an error on a missing date, so we drop the cases
without a date of onset. With gaps between the dates, **EpiNow2** 1.9.0 stopped
with an error on these data. So `fill_missing()` adds the days without cases,
with a count of 0.

```{r epidemic_models_cases, eval=F}
## get incidence from onset dates
cases <- linelist %>%
  filter(!is.na(date_onset)) %>%
  group_by(date = date_onset) %>%
  summarise(confirm = n()) %>%
  ## add the days without cases, with a count of 0
  fill_missing(missing_dates = "zero")
```

We can then estimate R<sub>t</sub> using the `epinow` function. Some notes on
the inputs:

* The `generation_time` argument takes the generation time distribution inside
  the `gt_opts` function.
* We can provide any number of 'chained' delay distributions to the `delays`
  argument. We would simply add them to the `incubation_period` object with `+`
  within the `delay_opts` function.
* `horizon` in the `forecast_opts` function indicates how many days we want to
  project future incidence for.
* We pass additional options to the `stan` argument to specify how long
  we want to run the inference for. Increasing `samples` and `chains` will give
  you a more accurate estimate that better characterises uncertainty, however
  will take longer to run.

```{r epidemic_models_run_epinow, eval=F}
## run epinow
epinow_res <- epinow(
  data = cases,
  generation_time = gt_opts(generation_time),
  delays = delay_opts(incubation_period),
  forecast = forecast_opts(horizon = 21),
  stan = stan_opts(samples = 750, chains = 4)
)
```

#### Analysing outputs {.unnumbered}

This page does not run the fit. The results below come from a fit that is
stored in the **appliedepidata** package. An earlier version of **EpiNow2** made
that fit. The code above is the way to make such a fit with **EpiNow2** 1.9.

On a fit from **EpiNow2** 1.9, the `plot()` function draws a summary figure, and
the `summary()` function gives a summary table. These functions stop with an
error on the stored fit. So we load the tables of the stored fit and plot them
with **ggplot2**:


```{r out.width=c('100%', '100%'), fig.show='hold', fig.width = 12, fig.height = 9, message=F, warning=F }
## load the stored estimates and the case counts that the fit used
estimates <- appliedepidata::get_data(name = "epidemic_models_summarised_estimates")
observations <- appliedepidata::get_data(name = "epidemic_models_reported_cases")

## plot summary figure
ggplot(
  data = filter(estimates, variable %in% c("R", "reported_cases")),
  aes(x = date)
) +
  ## observed case counts, in the reported cases panel only
  geom_col(
    data = mutate(observations, variable = "reported_cases"),
    aes(y = confirm),
    fill = "grey70",
    width = 1
  ) +
  ## 90% credible interval and median, coloured by estimate type
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90, fill = type), alpha = 0.3) +
  geom_line(aes(y = median, colour = type)) +
  ## use label_parsed to allow subscript label
  facet_wrap(
    ~ variable,
    ncol = 1,
    scales = "free_y",
    labeller = as_labeller(c("R" = "R[t]", "reported_cases" = "Reported~cases"), label_parsed),
    strip.position = 'left'
  ) +
  labs(x = NULL, y = NULL, colour = NULL, fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    strip.background = element_blank(),
    strip.placement = 'outside',
    legend.position = 'bottom'
  )
```

We can also look at various summary statistics:

```{r epidemic_models_epinow_summary,}
## summary table of the stored fit
summary_table <- appliedepidata::get_data(name = "epidemic_models_summary")
summary_table[, c("measure", "estimate")]
```

For further analyses and custom plotting, you can access the summarised daily
estimates of a fit from **EpiNow2** 1.9 with `summary(epinow_res, type = "parameters")`.
We will convert this from the default `data.table` to a `tibble` for ease of use
with **dplyr**. The table below shows the same estimates from the stored fit.

```{r epidemic_models_to_tibble, eval=F}
## extract summary and convert to tibble
estimates <- as_tibble(summary(epinow_res, type = "parameters"))
estimates
```

```{r epidemic_models_tibble_show,  echo = F}
## show outputs
estimates <- as_tibble(estimates)
DT::datatable(
  estimates,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 5, scrollX=T), class = 'white-space: nowrap'
)
```

As an example, let's make a plot of the doubling time and R<sub>t</sub>. We will
only look at the first few months of the outbreak when R<sub>t</sub> is well
above one, to avoid plotting extremely high doublings times.

We use the formula `log(2)/growth_rate` to calculate the doubling time from the
estimated growth rate.

```{r epidemic_models_plot_epinow_cusotom, out.width=c('100%', '100%'), fig.show='hold', fig.width = 12, fig.height = 9, message=F, warning=F}

## make wide df for median plotting
df_wide <- estimates %>%
  filter(
    variable %in% c("growth_rate", "R"),
    date < as.Date("2014-09-01")
  ) %>%
  ## convert growth rates to doubling times
  mutate(
    across(
      c(median, lower_90:upper_90),
      ~ case_when(
        variable == "growth_rate" ~ log(2)/.x,
        TRUE ~ .x
      )
    ),
    ## rename variable to reflect transformation
    variable = replace(variable, variable == "growth_rate", "doubling_time")
  )

## make long df for quantile plotting
df_long <- df_wide %>%
  ## here we match matching quantiles (e.g. lower_90 to upper_90)
  pivot_longer(
    lower_90:upper_90,
    names_to = c(".value", "quantile"),
    names_pattern = "(.+)_(.+)"
  )

## make plot
ggplot() +
  geom_ribbon(
    data = df_long,
    aes(x = date, ymin = lower, ymax = upper, alpha = quantile),
    color = NA
  ) +
  geom_line(
    data = df_wide,
    aes(x = date, y = median)
  ) +
  ## use label_parsed to allow subscript label
  facet_wrap(
    ~ variable,
    ncol = 1,
    scales = "free_y",
    labeller = as_labeller(c(R = "R[t]", doubling_time = "Doubling~time"), label_parsed),
    strip.position = 'left'
  ) +
  ## manually define quantile transparency
  scale_alpha_manual(
    values = c(`20` = 0.7, `50` = 0.4, `90` = 0.2),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    x = NULL,
    y = NULL,
    alpha = "Credible\ninterval"
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %d\n%Y"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.background = element_blank(),
    strip.placement = 'outside'
  )

```

<!-- ======================================================= -->
### EpiEstim {.unnumbered}

To run **EpiEstim**, we need to provide data on daily incidence and specify the
serial interval (i.e. the distribution of delays between symptom onset of
primary and secondary cases). 

Incidence data can be provided to **EpiEstim** as a vector, a data frame, or an `incidence`
object from the original **incidence** package. You can even distinguish between imports
and locally acquired infections; see the documentation at `?estimate_R` for
further details.  

We will create the input using **incidence2**. See the page on [Epidemic curves](epicurves.qmd) for more examples with the **incidence2** package. Since there have been updates to the **incidence2** package that don't completely align with `estimate_R()`'s expected input, there are some minor additional steps needed. The incidence object consists of a tibble with dates and their respective case counts. We use `complete()` from **tidyr** to ensure all dates are included (even those with no cases), and then `rename()` the columns to align with what is expected by `estimate_R()` in a later step.  

```{r epidemic_models_epiestim_incidence,}
## get incidence from onset date
cases <- incidence2::incidence(linelist, date_index = "date_onset") %>% # get case counts by day
  tidyr::complete(date_index = seq.Date(                              # ensure all dates are represented
    from = min(date_index, na.rm = T),
    to = max(date_index, na.rm=T),
    by = "day"),
    fill = list(count = 0)) %>%                                       # convert NA counts to 0
  rename(I = count,                                                   # rename to names expected by estimate_R()
         dates = date_index)
```

The package provides several options for specifying the serial interval, the
details of which are provided in the documentation at `?estimate_R`. We will
cover two of them here.

#### Using serial interval estimates from the literature {.unnumbered}

Using the option `method = "parametric_si"`, we can manually specify the mean and
standard deviation of the serial interval in a `config` object created using the
function `make_config`. We use a mean and standard deviation of 12.0 and 5.2, respectively, defined in
[this paper](https://bmcmedicine.biomedcentral.com/articles/10.1186/s12916-014-0196-0).

By default, `estimate_R()` uses weekly sliding windows from the second day of the
data. **EpiEstim** warns that R<sub>t</sub> estimated from very few cases is
unreliable: with its default settings, it needs at least 11 cases in a window.
The first weeks of this outbreak hold fewer. So we start the weekly windows at the
first 7-day window that holds at least 11 cases. Both runs below use these
windows:

```{r epidemic_models_epiestim_config,}
## daily case counts, one row per day
daily_cases <- cases$I

## number of cases in the 7-day window that starts on each day
window_cases <- sapply(
  seq_len(length(daily_cases) - 6),
  function(t) sum(daily_cases[t:(t + 6)])
)

## weekly sliding windows from the first window with at least 11 cases
weekly_start <- max(2, which(window_cases >= 11)[1]):(length(daily_cases) - 6)
weekly_end <- weekly_start + 6

## make config
config_lit <- make_config(
  mean_si = 12.0,
  std_si = 5.2,
  t_start = weekly_start,
  t_end = weekly_end
)
```

We can then estimate R<sub>t</sub> with the `estimate_R` function:

```{r epidemic_models_epiestim_lit,  warning = FALSE}

cases <- cases %>% 
     filter(!is.na(dates))


#create a dataframe for the function estimate_R()
cases_incidence <- data.frame(dates = seq.Date(from = min(cases$dates),
                               to = max(cases$dates), 
                               by = 1))

cases_incidence <- left_join(cases_incidence, cases, by = "dates") %>% 
     select(dates, I) %>% 
     mutate(I = ifelse(is.na(I), 0, I))

epiestim_res_lit <- estimate_R(
  incid = cases_incidence,
  method = "parametric_si",
  config = config_lit
)
```

and plot the R<sub>t</sub> estimates. The `plot()` function of **EpiEstim**
raises a deprecation warning from **ggplot2**, so we write a short function that
plots the estimates with **ggplot2**. It dates each estimate by the end of its
time window:

```{r epidemic_models_epiestim_lit_plot,  warning = FALSE}
## plot the mean R estimate and its 95% credible interval
plot_rt <- function(res) {
  res$R %>%
    ## drop the time windows without an estimate
    filter(!is.na(`Mean(R)`)) %>%
    ## date each estimate by the end of its time window
    mutate(date = res$dates[t_end]) %>%
    ggplot(aes(x = date)) +
    geom_ribbon(aes(ymin = `Quantile.0.025(R)`, ymax = `Quantile.0.975(R)`), alpha = 0.3) +
    geom_line(aes(y = `Mean(R)`)) +
    geom_hline(yintercept = 1, linetype = 2) +
    labs(x = NULL, y = expression(R[t])) +
    theme_minimal(base_size = 14)
}

plot_rt(epiestim_res_lit)
```

#### Using serial interval estimates from the data {.unnumbered}

As we have data on dates of symptom onset _and_ transmission links, we can
also estimate the serial interval from the linelist by calculating the delay
between onset dates of infector-infectee pairs. As we did in the **EpiNow2**
section, we will use the `get_pairwise` function from the **epicontacts**
package, which allows us to calculate pairwise differences of linelist
properties between transmission pairs. We first create an epicontacts object
(see [Transmission chains](transmission_chains.qmd) page for further details):

```{r epidemic_models_epicontacts_epiestim, eval=F}
## generate contacts
contacts <- linelist %>%
  transmute(
    from = infector,
    to = case_id
  ) %>%
  drop_na()

## generate epicontacts object
epic <- make_epicontacts(
  linelist = linelist,
  contacts = contacts, 
  directed = TRUE
)
```

We then fit the difference in onset dates between transmission pairs, calculated
using `get_pairwise`, to a gamma distribution. We use the handy `fit_disc_gamma`
from the **epitrix** package for this fitting procedure, as we require a
_discretised_ distribution.

```{r epidemic_models_incubation_estimate_epiestim,  warning = FALSE}
## estimate gamma serial interval, without the pairs that have no onset date
serial_interval <- fit_disc_gamma(na.omit(get_pairwise(epic, "date_onset")))
```

We then pass this information to the `config` object, run **EpiEstim**
again and plot the results:

```{r epidemic_models_epiestim_emp,  warning = FALSE}
## make config
config_emp <- make_config(
  mean_si = serial_interval$mu,
  std_si = serial_interval$sd,
  t_start = weekly_start,
  t_end = weekly_end
)

## run epiestim
epiestim_res_emp <- estimate_R(
  incid = cases_incidence,
  method = "parametric_si",
  config = config_emp
)

## plot outputs
plot_rt(epiestim_res_emp)
```

#### Specifying estimation time windows {.unnumbered}

You can also choose the start of the windows yourself, for example to estimate
R<sub>t</sub> only from 1 June 2014, as shown below. Unfortunately, **EpiEstim** only provides a very
clunky way of specifying these estimations times, in that you have to provide a
vector of __integers__ referring to the start and end dates for each time
window.

```{r epidemic_models_epiestim_config_late,}

## define a vector of dates starting on June 1st
start_dates <- seq.Date(
  as.Date("2014-06-01"),
  max(cases$dates) - 7,
  by = 1
) %>%
  ## subtract the starting date to convert to numeric
  `-`(min(cases$dates)) %>%
  ## convert to integer
  as.integer() %>%
  ## add 1, because day 1 of the incidence data is its first date
  `+`(1L)

## add six days for a one week sliding window
end_dates <- start_dates + 6
  
## make config
config_partial <- make_config(
  mean_si = 12.0,
  std_si = 5.2,
  t_start = start_dates,
  t_end = end_dates
)
```
Now we re-run **EpiEstim** and can see that the estimates only start from June:

```{r epidemic_models_epiestim_config_late_run,}

## run epiestim
epiestim_res_partial <- estimate_R(
  incid = cases_incidence,
  method = "parametric_si",
  config = config_partial
)

## plot outputs
plot_rt(epiestim_res_partial)

```

#### Analysing outputs {.unnumbered}

The main outputs can be accessed via `$R`. As an example, we will create a plot of
R<sub>t</sub> and a measure of "transmission potential" given by the product of
R<sub>t</sub> and the number of cases reported on that day; this represents the
expected number of cases in the next generation of infection.

```{r epidemic_models_epiestim_plot_full, out.width=c('100%', '100%'), fig.show='hold', fig.width = 12, fig.height = 9, message=F, warning=F}

## make wide dataframe for median
df_wide <- epiestim_res_lit$R %>%
  rename_all(clean_labels) %>%
  rename(
    lower_95_r = quantile_0_025_r,
    lower_90_r = quantile_0_05_r,
    lower_50_r = quantile_0_25_r,
    upper_50_r = quantile_0_75_r,
    upper_90_r = quantile_0_95_r,
    upper_95_r = quantile_0_975_r,
    ) %>%
  ## drop the time windows without an estimate
  filter(!is.na(mean_r)) %>%
  mutate(
    ## extract the median date from t_start and t_end
    dates = epiestim_res_emp$dates[round(map2_dbl(t_start, t_end, median))],
    var = "R[t]"
  ) %>%
  ## merge in daily incidence data
  left_join(cases, "dates") %>%
  ## calculate risk across all r estimates
  mutate(
    across(
      lower_95_r:upper_95_r,
      ~ .x*I,
      .names = "{str_replace(.col, '_r', '_risk')}"
    )
  ) %>%
  ## seperate r estimates and risk estimates
  pivot_longer(
    contains("median"),
    names_to = c(".value", "variable"),
    names_pattern = "(.+)_(.+)"
  ) %>%
  ## assign factor levels
  mutate(variable = factor(variable, c("risk", "r")))

## make long dataframe from quantiles
df_long <- df_wide %>%
  select(-variable, -median) %>%
  ## seperate r/risk estimates and quantile levels
  pivot_longer(
    contains(c("lower", "upper")),
    names_to = c(".value", "quantile", "variable"),
    names_pattern = "(.+)_(.+)_(.+)"
  ) %>%
  mutate(variable = factor(variable, c("risk", "r")))

## make plot
ggplot() +
  geom_ribbon(
    data = df_long,
    aes(x = dates, ymin = lower, ymax = upper, alpha = quantile),
    color = NA
  ) +
  geom_line(
    data = df_wide,
    aes(x = dates, y = median),
    alpha = 0.2
  ) +
  ## use label_parsed to allow subscript label
  facet_wrap(
    ~ variable,
    ncol = 1,
    scales = "free_y",
    labeller = as_labeller(c(r = "R[t]", risk = "Transmission~potential"), label_parsed),
    strip.position = 'left'
  ) +
  ## manually define quantile transparency
  scale_alpha_manual(
    values = c(`50` = 0.7, `90` = 0.4, `95` = 0.2),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    x = NULL,
    y = NULL,
    alpha = "Credible\ninterval"
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %d\n%Y"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.background = element_blank(),
    strip.placement = 'outside'
  )
  
```

<!-- ======================================================= -->
## Projecting incidence {  }

### EpiNow2 {.unnumbered}

Besides estimating R<sub>t</sub>, **EpiNow2** also supports forecasting of
R<sub>t</sub> and projections of case numbers. All you need to do is give
`forecast_opts(horizon = 21)` to the `forecast` argument of your `epinow`
function call. The `horizon` is the number of days you want to project into the
future. See the **EpiNow2** section under the "Estimating R<sub>t</sub>" for
details on how to get **EpiNow2** up and running. In this section, we will just
plot the forecast from the stored fit.

```{r epidemic_models_episoon, out.width=c('100%', '100%'), fig.show='hold', fig.width = 12, fig.height = 9, message=F, warning=F}

## define minimum date for plot
min_date <- as.Date("2015-03-01")

## load the summarised estimates of the stored fit
estimates <- as_tibble(appliedepidata::get_data(name = "epidemic_models_summarised_estimates"))

## load the case counts that the stored fit used
observations <- as_tibble(appliedepidata::get_data(name = "epidemic_models_reported_cases")) %>%
  filter(date > min_date)

## extract forecasted estimates of case numbers
df_wide <- estimates %>%
  filter(
    variable == "reported_cases",
    type == "forecast",
    date > min_date
  )

## convert to even longer format for quantile plotting
df_long <- df_wide %>%
  ## here we match matching quantiles (e.g. lower_90 to upper_90)
  pivot_longer(
    lower_90:upper_90,
    names_to = c(".value", "quantile"),
    names_pattern = "(.+)_(.+)"
  )

## make plot
ggplot() +
  geom_col(
    data = observations,
    aes(x = date, y = confirm),
    width = 1
  ) +
  geom_ribbon(
    data = df_long,
    aes(x = date, ymin = lower, ymax = upper, alpha = quantile),
    color = NA
  ) +
  geom_line(
    data = df_wide,
    aes(x = date, y = median)
  ) +
  geom_vline(xintercept = min(df_long$date), linetype = 2) +
  ## manually define quantile transparency
  scale_alpha_manual(
    values = c(`20` = 0.7, `50` = 0.4, `90` = 0.2),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    x = NULL,
    y = "Daily reported cases",
    alpha = "Credible\ninterval"
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %d\n%Y"
  ) +
  theme_minimal(base_size = 14)

```

### projections {.unnumbered}

The **projections** package developed by RECON makes it very easy to make short
term incidence forecasts, requiring only knowledge of the effective reproduction
number R<sub>t</sub> and the serial interval. Here we will cover how to use
serial interval estimates from the literature and how to use our own estimates
from the linelist.

#### Using serial interval estimates from the literature {.unnumbered}

**projections** requires a discretised serial interval distribution of the class
`distcrete` from the package **distcrete**. We will use a gamma distribution
with a mean of 12.0 and and standard deviation of 5.2 defined in
[this paper](https://bmcmedicine.biomedcentral.com/articles/10.1186/s12916-014-0196-0). To
convert these values into the shape and scale parameters required for a gamma
distribution, we will use the function `gamma_mucv2shapescale` from the
**epitrix** package.

```{r epidemic_models_projections_distcrete,}

## get shape and scale parameters from the mean mu and the coefficient of
## variation (e.g. the ratio of the standard deviation to the mean)
shapescale <- epitrix::gamma_mucv2shapescale(mu = 12.0, cv = 5.2/12)

## make distcrete object
serial_interval_lit <- distcrete::distcrete(
  name = "gamma",
  interval = 1,
  shape = shapescale$shape,
  scale = shapescale$scale
)

```

Here is a quick check to make sure the serial interval looks correct. We
access the density of the gamma distribution we have just defined by `$d`, which
is equivalent to calling `dgamma`:

```{r epidemic_models_projections_distcrete_plot,}

## check to make sure the serial interval looks correct
ggplot(
  data = tibble(x = 0:50, y = serial_interval_lit$d(0:50)),
  aes(x = x, y = y)
) +
  geom_area() +
  labs(x = "Serial interval", y = "Density")

```

#### Using serial interval estimates from the data {.unnumbered}

As we have data on dates of symptom onset _and_ transmission links, we can
also estimate the serial interval from the linelist by calculating the delay
between onset dates of infector-infectee pairs. As we did in the **EpiNow2**
section, we will use the `get_pairwise` function from the **epicontacts**
package, which allows us to calculate pairwise differences of linelist
properties between transmission pairs. We first create an epicontacts object
(see [Transmission chains](transmission_chains.qmd) page for further details):

```{r epidemic_models_epicontacts_projections, eval=F}
## generate contacts
contacts <- linelist %>%
  transmute(
    from = infector,
    to = case_id
  ) %>%
  drop_na()

## generate epicontacts object
epic <- make_epicontacts(
  linelist = linelist,
  contacts = contacts, 
  directed = TRUE
)
```

We then fit the difference in onset dates between transmission pairs, calculated
using `get_pairwise`, to a gamma distribution. We use the handy `fit_disc_gamma`
from the **epitrix** package for this fitting procedure, as we require a
_discretised_ distribution.

```{r epidemic_models_incubation_estimate_projections,  warning = FALSE}
## estimate gamma serial interval, without the pairs that have no onset date
serial_interval <- fit_disc_gamma(na.omit(get_pairwise(epic, "date_onset")))

## inspect estimate
serial_interval[c("mu", "sd")]
```

#### Projecting incidence {.unnumbered}

To project future incidence, we still need to provide historical incidence in
the form of an `incidence` object, as well as a sample of plausible
R<sub>t</sub> values. We will generate these values using the R<sub>t</sub>
estimates generated by **EpiEstim** in the previous section (under "Estimating
R<sub>t</sub>") and stored in the `epiestim_res_emp` object. In the code below,
we extract the mean and standard deviation estimates of R<sub>t</sub> for the
last time window of the outbreak (using the `tail` function to access the last
element in a vector), and simulate 1000 values from a gamma distribution using
`rgamma`. You can also provide your own vector of R<sub>t</sub> values that you
want to use for forward projections.

```{r epidemic_models_projection_setup,  warning = FALSE}

## create incidence object from dates of onset, without the cases that have no onset date
inc <- incidence::incidence(na.omit(linelist$date_onset))

## extract plausible r values from most recent estimate
mean_r <- tail(epiestim_res_emp$R$`Mean(R)`, 1)
sd_r <- tail(epiestim_res_emp$R$`Std(R)`, 1)
shapescale <- gamma_mucv2shapescale(mu = mean_r, cv = sd_r/mean_r)
plausible_r <- rgamma(1000, shape = shapescale$shape, scale = shapescale$scale)

## check distribution
ggplot(data = tibble(r = plausible_r), aes(x = r)) +
  geom_histogram(bins = 30) +
  labs(x = expression(R[t]), y = "Counts")

```

We then use the `project()` function to make the actual forecast. We specify how
many days we want to project for via the `n_days` arguments, and specify the
number of simulations using the `n_sim` argument.

```{r epidemic_models_make_projection,}

## make projection
proj <- project(
  x = inc,
  R = plausible_r,
  si = serial_interval$distribution,
  n_days = 21,
  n_sim = 1000
)

```

We can then plot the incidence with **ggplot2** and add the projections with the
`add_projections()` function. The `plot()` function of **incidence** raises a
deprecation warning from **ggplot2**, so we do not use it. We can easily subset
the `incidence` object to only show the most recent cases by using the square
bracket operator.

```{r epidemic_models_plot_projection, out.width=c('100%', '100%'), fig.show='hold', fig.width = 12, fig.height = 9, message=F, warning=F}

## plot the incidence
incidence_plot <- ggplot() +
  geom_col(
    data = as.data.frame(inc[inc$dates > as.Date("2015-03-01")]),
    aes(x = dates, y = counts),
    width = 1
  ) +
  theme_minimal(base_size = 14)

## add the projections
add_projections(incidence_plot, proj)

```

You can also easily extract the raw estimates of daily case numbers by
converting the output to a dataframe.

```{r epidemic_models_projection_df, eval=F, warning = FALSE}
## convert to data frame for raw data
proj_df <- as.data.frame(proj)
proj_df
```

```{r epidemic_models_projection_dt,  echo = F}

## convert to data frame for raw data
proj_df <- as.data.frame(proj)

## data table output
DT::datatable(
  proj_df[1:11],
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 10, scrollX=T), class = 'white-space: nowrap'
)

```


<!-- ======================================================= -->
## Resources {  }

* [Here is the paper](https://www.sciencedirect.com/science/article/pii/S1755436519300350) describing
  the methodology implemented in **EpiEstim**.
* [Here is the paper](https://wellcomeopenresearch.org/articles/5-112/v1) describing
  the methodology implemented in **EpiNow2**.
* [Here is a paper](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008409) describing
  various methodological and practical considerations for estimating R<sub>t</sub>.
